\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-08 chapters/wcble.tex
Proofread

\section{Introduction}
\label{sec:wcble/intro}
\todohsm{do'nt repeat in introduction, rather make line clear}

\Ac{eeg}-based visual oddball \acp{bci} establish communication with paralyzed
individuals by decoding \acp{erp} obtained from time-modulated flashing
stimuli (see Figure~\ref{fig:bci/loop}).
These \acp{erp} consist of multiple \ac{erp}-components, peaks and troughs in
\ac{erp} waveform (see Figure~\ref{fig:bci/erp}.
These components each have their own distinct amplitude and latency.
\Ac{erp} component latency can reflect a wide range of cognitive,
neurophysiological and clinical properties, that can be confounding factors in
the development of \acp{bci}~\cite{Luck2014}.
Latency is often assessed as the timing of an \ac{erp} component peak, obtained
after averaging over trials.
Latencies of non-averaged, single-trial \acp{erp} should capture even more
relevant information, yet they are very hard to measure due to low \ac{snr} of
single trial \acp{erp}.
Within-session variability of \ac{erp} latency, called \ac{erp} latency
\emph{jitter}, has been shown to influence \ac{bci}
accuracy~\cite{Thompson2012}.
This effect is especially prominent when bringing \ac{bci} development from the
lab to the clinical setting with patients in the target population.
\textcite{Zisk2021} showed that patients with \ac{als} have significantly
increased ERP jitter.
\textcite{Arico2014} show that latency jitter negatively affects \ac{bci}
performance, and disproportionally so when the user is not directly gazing at
intended targets, such as is the case for individuals with \ac{sspgi}.
Therefore, properly accounting for \ac{erp} latency jitter can contribute to
the development of gaze-independent visual \acp{bci}.

@@

The presence of latency jitter introduces multiple issues to \ac{erp} analysis, the
most prominent of which is the smearing effect as illustrated
in~\autoref{fig:wcble/smearing}. In \ac{erp} analysis, single-trial \acp{erp} are usually
averaged per condition to cope with the unfavorable \ac{snr}?
When averaging over multiple identical responses with jittered latencies, the shape
and amplitude of the average do not reflect the properties
of the original signal.
Because the amplitude of the average is lower, the \ac{snr} will also be negatively impacted.
More generally, the smearing effect leads to difficulties interpreting \ac{erp}
results and making inferences about amplitude or latency effects, because these
are entangled when using averaging~\cite{Luck2014}.

@@

The smearing effect equally impacts the \ac{snr} of information captured by a
classifier's parameters when training with a procedure that does not account for
this jitter, and thus also affects its performance~\cite{Thompson2012}.
As an illustrative example, think of an \ac{lda} \ac{erp} classifier that uses
class averages over \ac{erp} epochs to construct its between-class scatter
matrix.
These per-class averages will be affected by the smearing effect and will be
less effective templates for \ac{erp} retrieval.
Similarly, the noise model in \ac{lda}'s  assumes stationarity and subtracts
the class-wise averages from the signal to calculate
the within-class scatter matrix, which will then also be affected.
Hence, effective latency estimation and jitter compensation can contribute to higher
decoding performance.

\subsection{Benefits of latency estimation}
As we will see later, a common approach to counteract the smearing effect is to correct single-trial
\acp{erp} for jitter before averaging.
This can be done by estimating the latency of each
single-trial \ac{erp}, and aligning them before averaging.
Latency estimation has multiple advantages:
\begin{enumerate*}[label={(\arabic*)}]
	\item Latency-based \ac{erp} alignment can help reveal the true shape and
	properties like peak amplitude and latency of \acp{erp} across
	experimental conditions. These are variables of interest for
	testing electrophysiological an behavioral hypotheses.
	Additionally, latency estimation allows for the extraction of
	other descriptors from a set of single-trial \acp{erp}, like
	amplidute or latency variance~\cite{Hultsch2004}, which can
	also be of interest as studied variables.
  For instance, in the work of \textcite{Saville2014}, the inter-trial latency
  variability itself is determined as a correlate of working memory performance.
	\item Counteracting the smearing effect will boost the signal-to-noise
	ratio of averaged \acp{erp} by increasing the amplitude of the
	\ac{erp} signal. This in turn can reveal smaller or more jittered \ac{erp}
  components which would otherwise obscured or blurred due to desynchronization.
    These \emph{deblurred} templates can then be used for analysis of the \ac{erp}
	waveform, or as templates of the signal of interest in
	\acp{bci}~\cite{Arico2014}.
	%this enables decoders to take into account the relatively low amplitude
	%earlier components, leading to applications in settings where
	%information from other components than the P300 need to be exploited,
	%like covert visual attention paradigms~\cite{arico2014influence,
	%hardiansyah2020single}.\todo[inline]{rework this to latency features
	%because smaller visual components might not really be revealed
	%since they're already time-locked}
	\item Finally, the introduction of latency estimation and alignment generally
	improves \ac{bci} classification performance.
  Single-trial latencies can be	used as features to improve
  classification~\cite{Hardiansyah2020}.
	Furthermore, compensating for inter-trial differences, especially when
	extended to inter-session or inter-subject differences, can aid in transfer learning or
	generalization across subjects and protocols\cite{Iturrate2014}.
\end{enumerate*}

@@

The simplest method to estimate single-trial latency is
\emph{peak picking}
Here, the latency of an \ac{erp} is determined as the
time-point of it's maximum (or minimum) amplitude relative to stimulus
onset.
While straightforward, this method does not perform well in low
\ac{snr} conditions, unless combined with filtering.
Filtering can suppress the noise contaminating \ac{eeg} trials, lowering
the risk of picking a noise peak instead of a true peak of the \ac{erp}.
Several filtering approaches have been developed to be used in
conjuction with peak-picking, in the time domain~\cite{
  Carlton1980, McCarthy1981, Karjalainen1999, Nishida1999,  Sparacino2002,
  Georgiadis2005,	D’Avanzo2011}
and in the time-frequency domain~\cite{
  Quiroga2003, Wang2007,Hu2010}.
The aforementioned analysis and performance prediction method proposed
by \cite{Arico2014} estimates the single-trial P3 latencies of attended
epochs by P3 peak-picking after filtering in the time-frequency domain, and
corrects for jitter by temporally aligning all target epochs to the average
latency, i.e., by shifting the epochs in time such that the P3 peaks all fall
at the same time instance.
Filtering can
also be done in the spatial domain. The spatial filter can be constructed
using \ac{pca}~\cite{Ouyang2017}, ICA~\cite{Townsend1999, Jung2001,
	Milne2011}, or as a spatial LCMV beamformer~\cite{Treder2016}.

Template matching is generally preferred to peak picking.
The algorithm proposed by \textcite{Woody1967} in 1976 is a simple and elegant
latency estimation method that is still being used and is even
considered amongst the more performant techniques for
\acp{erp}~\cite{Thornton2008, Ouyang2017}.
It starts with an average \ac{erp} as a rough estimation of the aligned template
\ac{erp}, and iteratively refines this template by determining at each step the time point of maximum
cross-correlation of this template with all single-trial \acp{erp} and aligning them
on these time points to form the next template.
Similarily, cross-correlation template matching can be combined with filtering.
\cite{Souloumiac2013} uses XDAWN to construct spatial filter before
template matching the spatially filtered signal. \cite{Iturrate2014} uses a
spatiotemporal filter. Instead of aligning the
latencies to obtain the template for the next iteration, weighted averaging can
also be used~\cite{Gasser1983}.

Despite good performance, the iterative scheme of Woody's algorithm has a risk of converging to a local
optimum.
To avoid this, a fitness function can be defined to jointly optimize
the set of single-trial \ac{erp} latencies, which can be optimized with a genetic algorithm~\cite{Pelo2018}.
Alternative methods are based on graph
optimization~\cite{Dimitriadis2018}, hidden process models~\cite{Kim2020} or by fitting an \ac{erp} model with
maximum likelihood estimation~\cite{Gratton1989, Tuan1987, Moecks1988,Puce1994}.

\subsection{Multi-component approaches}
\label{sec:wcble/literature/multi-comp}
In the previous section, we have abstracted away the fact that \acp{erp} are
multi-component signals, which is not taking into account by the aforementioned
methods.
Multiple components can be time-locked to different events or neurophysiological
processes and therefore are not necessarily time-locked to each other. They
each can have different latency variabilities.
The presence of more than one component can hamper the
performance of some latency estimation algorithms, and it often only possible to extract only the
latencies of the largest component present~\cite{Ouyang2017}. When aligning trials
to one component and averaging, this will introduce blurring in the other
components~\cite{Ouyang2020}.

Most of the algorithms above can be adapted to work on multiple \ac{erp} components
by carefully selecting peaks or pre-determined regions of interest.
\cite{Hardiansyah2020} uses an \ac{snr} boosting method, applied to a specific
time-window for each \ac{erp} component to extract a latency per component.
Methods based on spatial decomposition filters like \ac{pca}, ICA or XDAWN can be leveraged
to utilize multiple of their components as filters, resulting in multiple time
series where peack picking or Woody's algorithm can be applied. In general,
however, these algorithms lack an integrated approach to deal with multiple
\ac{erp} components. A clear evaluation of the performance of these
methods in multi-component settings is lacking.

Some algorithms separate components into stimulus and response locked component clusters based on their
latency distributions~\cite{Jung2001, Takeda2008,	Zhang1998, Yin2009}.
These algorithms do, however, require a response
button trigger signal as input to determine the response time, which is
not always present or applicable in an \ac{erp} experiment, especially in
the case of \acp{bci}.
The RIDE algorithm~\cite{Ouyang2011, Ouyang2015,	Wang2015, Ouyang2016, Ouyang2020}
is able to do this without entirely relying on the response time, and can
separate stimulus-locked, response-locked, and multiple central component clusters.
Others, like spatial filtering methods that use decomposition methods to
determine the spatial filter can be be modified to separate
components~\cite{Ouyang2017}.

Finally, the smearing problem can also be overcome by algorithms that are not
based on \ac{erp} component latency estimation, but which use time warping-based methods
to enhance the shape of the averaged template \ac{erp}, like Dynamic Time
Warping~\cite{Gupta1996, Wang2001, Zoumpoulaki2015}, Correlation-Optimized
Warping~\cite{Skov2006}, or Fast Variational
Alignment~\cite{Flotho2021}.
These methods work
fundamentally different from the latency estimation methods, and cannot
directly be used to extract latency features.

@@

Most of these methods suffer from a common drawback: they cannot be used in a
decoding scheme to improve performance for incoming epochs with unknown
class labels that contain either a target or non-target \ac{erp} response.
They can be used offline on a set of labeled epochs for testing hypotheses
concerning latency and jitter, for aligning templates, or for \ac{bci} performance
prediction, but not for \ac{erp} classification.
While some of the aforementioned methods could be adapted to perform
classification tasks, few studies investigate how to exploit this latency estimation
for jitter-resistant decoding.
\cite{Hardiansyah2020} incorporated single-trial latencies in
classification by peak-picking within a given
\ac{erp} time window, unaware of the class of the epoch under investigation.
The \ac{cble} algorithm introduced
by \cite{Thompson2012} also explicitly applies latency estimation in a
decoding setting.
\cite{Thompson2012} initially formulated \ac{cble} as an off-line performance
prediction method.
Later, its output was successfully adapted to compensate for jitter to improve
decoder performance~\cite{Mowla2017,Zisk2022}.

Time series classification algorithms~\cite{Abanda2019}
that are robust to jitter can be used in a decoding setting,
but, in general, have scarcely been applied to \ac{erp} decoding.
Data augmentation involving jittering the training
data~\cite{Krell2018,Zisk2022} and Riemannian Geometry methods using spatial
covariances as features~\cite{Aydarkhanov2020} have both been shown
to perform well in the presence of \ac{erp} jitter.
In this work, we opted to apply \ac{cble} because it has successfully been applied to
classify jittered \acs{erp}.
We adapt the \ac{cble} to an iterative method akin to Woody's scheme that can better
compensate for jitter, to improve covert VSA decoding performance.

@@

Consider a training set of $N$ EEG epochs with $C$ channels of $S$
samples $\{\mat{X}_n^\mathrm{train} \in \mathbb{R}^{C\times S}\}_{n=1}^N$
with corresponding training labels $\mathbf{l^\mathrm{train}} \in \{\mathrm{target},
	\textrm{non-target}\}_{n=1}^N$, and a
similar testing set of $M$ epochs $\{\mat{X}_m^\mathrm{test} \in
	\mathbb{R}^{C\times S}\}_{m=1}^M$.
We assume that the sampling period is $T$, i.e. that the sample with index $s$ is sampled at time $sT$.
In the following, we use the matrix slicing notation to denote row or column intervals extracted from a matrix.
For instance, $\mat{X}[:,s_1:s_2]$ denotes all columns of $\mat{X}$ with indices between $s_1$ (included) and $s_2$ (excluded).

\Ac{cble}, summarized in \autoref{sec:wcble/app/cble-alg}, works by training a
\textit{first-stage} classifier $\mathcal{C}(\theta,f)$
defined within a time period $[s_1:s_2]$ with a set of parameters $\theta$ and a
decision function $f(\mat{X}[:,s_1:s_2],\theta) \rightarrow
y$ outputting a classification score $y\in\mathbb{R}$ for a given epoch $\mat{X}$,
such that
\begin{equation}
  \theta = \mathrm{train}_\mathcal{C}(\{\mat{X}_n^\mathrm{train}[:,s_1:s_2]\}_{n=1}^N,\mathbf{l}^\mathrm{train})
  \label{eq:cble-train}
\end{equation}
Then,
%instead of applying the classifier directly to an unseen, cropped
%testing epoch $\mat{X}[:,s_1:s_2]$ and obtaining a single score value $y$,
$f$ can be applied to all (possibly overlapping) slices of length $s_2-s_1$ of
an epoch $\mat{X}$, resulting in a vector of score values
$\mathbf{y}=[y_1\,\ldots\,y_R]^T \in\mathbb{R}^R$ such that
\begin{equation}
  y_s = f(\mat{X}[:,s:s+(s_2-s_1)],\theta)\quad \forall s\in 1,\ldots,R
	\label{eq:score}
\end{equation}
with $R = S-(s_2-s_1)$.
To leverage \ac{cble} for \ac{erp} classification, the score vectors $\mathbf{y}$ can be
arranged in matrices $\mat{Y}^\mathrm{train}\in\mathbb{R}^{N\times R}$ and $\mat{Y}^\mathrm{test}\in\mathbb{R}^{M\times R}$.
These can be further classified by training a \textit{second-stage} classifier on
$\mat{Y}^\mathrm{train}$ and class labels $\mathbf{l}^\mathrm{train}$.
However, the resulting score-over-time vectors per epoch still suffer from jitter.
For classification, we follow the approach of \cite{Mowla2017}, using a
maximum-level hierarchical Daubechies-4 wavelet transform to reduce
dimensionality before classification with the second-stage classifier.
In the \ac{cble}-decoder, it is this wavelet transform that decreases the sensitivity
to latency differences, actively compensating for \ac{erp} latency jitter.

When using a simple spatiotemporal linear classifier as first-stage classifier,
\ac{cble} is equivalent to the first iteration of Woody's algorithm with the
spatiotemporal classifier weights as template.
\cite{Thompson2012}, \cite{Mowla2017} and \cite{Mowla2020} show
that \ac{cble} is relatively independent of the first-stage classifier for \ac{bci}
accuracy prediction and for \ac{erp} classification.
Therefore, we opt to use the variant of Linear Discriminant Analysis with block-Toeplitz
regularized covariance matrix (tLDA) proposed by \cite{Sosulski2022}, as the first-stage
classifier and logistic regression as second stage.

\subsection{Robust latency features}
\label{sec:robust-latency}
The \ac{erp} decoding method based on \ac{cble} introduced by \cite{Mowla2017} does
not make use of the extracted latencies, only passing score matrix $\mat{Y}$ on
to the second-stage classifier.
Furthermore, previous \ac{cble} works~\cite{Thompson2012,Mowla2020} only used these
latencies to correlate them with neurophysiological processes or to predict
decoder performance.
We noticed that, while \ac{cble} performance was unaffected, the classification
performance of our proposed method can be improved if the estimated
latencies are also made available as features to the second-stage classifier,
after a square transform for linear separability~\cite{Thompson2012}.

However, including these latency features give rise to the following issue when
classifying unseen data.
The \ac{cble} latency estimate is defined only for target epochs as
$s_\mathrm{target}=\argmax_s{y_s}$.
This is the point in time where the target class reaches largest separation
from the background noise and non-target class, indicating the target \ac{erp} is
most likely to occur here.
However, in a classifier test phase, it is not known a priori whether an
unseen epoch is a target or a non-target epoch.
This problem is solved by defining an estimated latency per class
$s_\mathrm{target}$ and $s_\mathrm{non-target}$ for every epoch,
regardless of its actual class.
The estimated class latencies can then be used as features  for training and
testing the second-stage classifier.
This way, the latencies of the testing data can be presented to the second-stage
classifier without knowledge of the testing data class labels, making them
useful in decoding.

In a similar manner to the target latency, the non-target latency could be defined
as~$s_\mathrm{non-target}=\argmin_s{y_s}$.
However, this is problematic since it is not evident how to estimate
latency of e.g. a P3 \ac{erp} component for a non-target epoch, since the
non-target class is characterized by the absence of this component.
\footnote{Depending on the stimulation and experimental design, \ac{erp} components can be present in one experimental
	condition or class and missing in another, or appear in multiple classes
	at different amplitudes. In the latter case, it could be possible
	to estimate the latency of an \ac{erp} in the conditions where it appears, but if
	its amplitude is lower or negligible in some conditions, latency estimates
	will be less accurate, hence this case is still problematic.}.
In fact, $y$ can have multiple local minima or entirely lack distinct peaks for
non-targets, rendering the minimum estimate meaningless.

>>>>

\section{Introduction}
\label{sec:wcble/intro}

\Ac{eeg}-based visual oddball \acp{bci} establish communication with paralyzed
individuals by decoding \acp{erp} obtained from time-modulated flashing
stimuli (see~\cref{fig:bci/loop}).
These \acp{erp} consist of multiple \ac{erp}-components, peaks, and troughs in
\ac{erp} waveform (see \cref{fig:bci/erp}).
These components each have their own distinct amplitude and latency.
\Ac{erp} component latency can reflect a wide range of cognitive,
neurophysiological, and clinical properties, that can be confounding factors in
the development of \acp{bci}~\cite{Luck2014}.
Latency is often assessed as the timing of an \ac{erp} component peak, obtained
after averaging over trials.
Latencies of non-averaged, single-trial \acp{erp} should capture even more
relevant information, yet they are very hard to measure due to the low \ac{snr} of
single-trial \acp{erp}.
Within-session variability of \ac{erp} latency, called \ac{erp} latency
\emph{jitter}, has been shown to influence \ac{bci}
accuracy~\cite{Thompson2012}.
This effect is especially prominent when bringing \ac{bci} development from the
lab to the clinical setting with patients in the target population.
\textcite{Zisk2021} showed that patients with \ac{als} have significantly
increased \ac{erp} jitter.
\textcite{Arico2014} showed that latency jitter negatively affects \ac{bci}
performance, and disproportionately so when the user is not directly gazing at
intended targets, such as is the case for individuals with \ac{sspgi}.
Therefore, properly accounting for \ac{erp} latency jitter can contribute to
the development of gaze-independent visual \acp{bci}.

@@

The presence of latency jitter introduces multiple issues to \ac{erp} analysis, the
most prominent of which is the smearing effect, as illustrated
in~\cref{fig:wcble/smearing}. In \ac{erp} analysis, single-trial \acp{erp} are usually
averaged per condition to cope with the unfavorable \ac{snr}.
When averaging over multiple identical responses with jittered latencies, the shape
and amplitude of the average do not reflect the properties
of the original signal.
Because the amplitude of the average is lower, the \ac{snr} will also be negatively impacted.
More generally, the smearing effect leads to difficulties interpreting \ac{erp}
results and making inferences about amplitude or latency effects because these
are entangled when using averaging~\cite{Luck2014}.

@@

The smearing effect equally impacts the \ac{snr} of information captured by a
classifier's parameters when training with a procedure that does not account for
this jitter, and thus also affects its performance~\cite{Thompson2012}.
As an illustrative example, think of an \ac{lda} \ac{erp} classifier that uses
class averages over \ac{erp} epochs to construct its between-class scatter
matrix.
These per-class averages will be affected by the smearing effect and will be
less effective templates for \ac{erp} retrieval.
Similarly, the noise model in \ac{lda} assumes stationarity and subtracts
the class-wise averages from the signal to calculate
the within-class scatter matrix, which will then also be affected.
Hence, effective latency estimation and jitter compensation can contribute to higher
decoding performance.

\subsection{Benefits of latency estimation}
As we will see later, a common approach to counteract the smearing effect is to correct single-trial
\acp{erp} for jitter before averaging.
This can be done by estimating the latency of each
single-trial \ac{erp} and aligning them before averaging.
Latency estimation has multiple advantages:
\begin{enumerate*}[label={(\arabic*)}]
	\item Latency-based \ac{erp} alignment can help reveal the true shape and
	properties like peak amplitude and latency of \acp{erp} across
	experimental conditions. These are variables of interest for
	testing electrophysiological and behavioral hypotheses.
	Additionally, latency estimation allows for the extraction of
	other descriptors from a set of single-trial \acp{erp}, like
	amplitude or latency variance~\cite{Hultsch2004}, which can
	also be of interest as studied variables.
  For instance, in the work of \textcite{Saville2014}, the inter-trial latency
  variability itself is determined as a correlate of working memory performance.
	\item Counteracting the smearing effect will boost the signal-to-noise
	ratio of averaged \acp{erp} by increasing the amplitude of the
	\ac{erp} signal. This in turn can reveal smaller or more jittered \ac{erp}
  components which would otherwise be obscured or blurred due to desynchronization.
    These \emph{deblurred} templates can then be used for analysis of the \ac{erp}
	waveform, or as templates of the signal of interest in
	\acp{bci}~\cite{Arico2014}.
	%this enables decoders to take into account the relatively low amplitude
	%earlier components, leading to applications in settings where
	%information from other components than the P300 need to be exploited,
	%like covert visual attention paradigms~\cite{arico2014influence,
	%hardiansyah2020single}.\todo[inline]{rework this to latency features
	%because smaller visual components might not really be revealed
	%since they're already time-locked}
	\item Finally, the introduction of latency estimation and alignment generally
	improves \ac{bci} classification performance.
  Single-trial latencies can be used as features to improve
  classification~\cite{Hardiansyah2020}.
	Furthermore, compensating for inter-trial differences, especially when
	extended to inter-session or inter-subject differences, can aid in transfer learning or
	generalization across subjects and protocols~\cite{Iturrate2014}.
\end{enumerate*}

@@

The simplest method to estimate single-trial latency is
\emph{peak picking}.
Here, the latency of an \ac{erp} is determined as the
time-point of its maximum (or minimum) amplitude relative to stimulus
onset.
While straightforward, this method does not perform well in low
\ac{snr} conditions, unless combined with filtering.
Filtering can suppress the noise contaminating \ac{eeg} trials, lowering
the risk of picking a noise peak instead of a true peak of the \ac{erp}.
Several filtering approaches have been developed to be used in
conjunction with peak-picking, in the time domain~\cite{
  Carlton1980, McCarthy1981, Karjalainen1999, Nishida1999, Sparacino2002,
  Georgiadis2005, D’Avanzo2011}
and in the time-frequency domain~\cite{
  Quiroga2003, Wang2007, Hu2010}.
The aforementioned analysis and performance prediction method proposed
by \textcite{Arico2014} estimates the single-trial P3 latencies of attended
epochs by P3 peak-picking after filtering in the time-frequency domain and
corrects for jitter by temporally aligning all target epochs to the average
latency, i.e., by shifting the epochs in time such that the P3 peaks all fall
at the same time instance.
Filtering can
also be done in the spatial domain. The spatial filter can be constructed
using \ac{pca}~\cite{Ouyang2017}, \ac{ica}~\cite{Townsend1999, Jung2001,
	Milne2011}, or as a spatial LCMV beamformer~\cite{Treder2016}.

Template matching is generally preferred to peak picking.
The algorithm proposed by \textcite{Woody1967} in 1976 is a simple and elegant
latency estimation method that is still being used and is even
considered among the more performant techniques for
\acp{erp}~\cite{Thornton2008, Ouyang2017}.
It starts with an average \ac{erp} as a rough estimation of the aligned template
\ac{erp}, and iteratively refines this template by determining at each step the time point of maximum
cross-correlation of this template with all single-trial \acp{erp} and aligning them
on these time points to form the next template.
Similarly, cross-correlation template matching can be combined with filtering.
\textcite{Souloumiac2013} uses XDAWN to construct a spatial filter before
template matching the spatially filtered signal. \textcite{Iturrate2014} uses a
spatiotemporal filter. Instead of aligning the
latencies to obtain the template for the next iteration, weighted averaging can
also be used~\cite{Gasser1983}.

Despite good performance, the iterative scheme of Woody's algorithm has a risk of converging to a local
optimum.
To avoid this, a fitness function can be defined to jointly optimize
the set of single-trial \ac{erp} latencies, which can be optimized with a genetic algorithm~\cite{Pelo2018}.
Alternative methods are based on graph
optimization~\cite{Dimitriadis2018}, hidden process models~\cite{Kim2020}, or by fitting an \ac{erp} model with
maximum likelihood estimation~\cite{Gratton1989, Tuan1987, Moecks1988, Puce1994}.

\subsection{Multi-component approaches}
\label{sec:wcble/literature/multi-comp}
In the previous section, we have abstracted away the fact that \acp{erp} are
multi-component signals, which is not taken into account by the aforementioned
methods.
Multiple components can be time-locked to different events or neurophysiological
processes and therefore are not necessarily time-locked to each other. They
each can have different latency variabilities.
The presence of more than one component can hamper the
performance of some latency estimation algorithms, and it is often only possible to extract the latencies of the largest component present~\cite{Ouyang2017}. When aligning trials
to one component and averaging, this will introduce blurring in the other
components~\cite{Ouyang2020}.

Most of the algorithms above can be adapted to work on multiple \ac{erp} components
by carefully selecting peaks or pre-determined regions of interest.
\textcite{Hardiansyah2020} uses an \ac{snr} boosting method, applied to a specific
time-window for each \ac{erp} component to extract a latency per component.
Methods based on spatial decomposition filters like \ac{pca}, \ac{ica}, or XDAWN can be leveraged
to utilize multiple of their components as filters, resulting in multiple time
series where peak picking or Woody's algorithm can be applied. In general,
however, these algorithms lack an integrated approach to deal with multiple
\ac{erp} components. A clear evaluation of the performance of these
methods in multi-component settings is lacking.

Some algorithms separate components into stimulus- and response-locked component clusters based on their
latency distributions~\cite{Jung2001, Takeda2008, Zhang1998, Yin2009}.
These algorithms do, however, require a response
button trigger signal as input to determine the response time, which is
not always present or applicable in an \ac{erp} experiment, especially in
the case of \acp{bci}.
The RIDE algorithm~\cite{Ouyang2011, Ouyang2015, Wang2015, Ouyang2016, Ouyang2020}
is able to do this without entirely relying on the response time and can
separate stimulus-locked, response-locked, and multiple central component clusters.
Others, like spatial filtering methods that use decomposition methods to
determine the spatial filter, can be modified to separate
components~\cite{Ouyang2017}.

Finally, the smearing problem can also be overcome by algorithms that are not
based on \ac{erp} component latency estimation but use time-warping-based methods
to enhance the shape of the averaged template \ac{erp}, like Dynamic Time
Warping~\cite{Gupta1996, Wang2001, Zoumpoulaki2015}, Correlation-Optimized
Warping~\cite{Skov2006}, or Fast Variational
Alignment~\cite{Flotho2021}.
These methods work
fundamentally differently from the latency estimation methods, and cannot
directly be used to extract latency features.

@@

Most of these methods suffer from a common drawback: they cannot be used in a
decoding scheme to improve performance for incoming epochs with unknown
class labels that contain either a target or non-target \ac{erp} response.
They can be used offline on a set of labeled epochs for testing hypotheses
concerning latency and jitter, for aligning templates, or for \ac{bci} performance
prediction, but not for \ac{erp} classification.
While some of the aforementioned methods could be adapted to perform
classification tasks, few studies investigate how to exploit this latency estimation
for jitter-resistant decoding.
\textcite{Hardiansyah2020} incorporated single-trial latencies in
classification by peak-picking within a given
\ac{erp} time window, unaware of the class of the epoch under investigation.
The \ac{cble} algorithm introduced
by \textcite{Thompson2012} also explicitly applies latency estimation in a
decoding setting.
\textcite{Thompson2012} initially formulated \ac{cble} as an offline performance
prediction method.
Later, its output was successfully adapted to compensate for jitter to improve
decoder performance~\cite{Mowla2017, Zisk2022}.

Time-series classification algorithms~\cite{Abanda2019}
that are robust to jitter can be used in a decoding setting,
but, in general, have scarcely been applied to \ac{erp} decoding.
Data augmentation involving jittering the training
data~\cite{Krell2018, Zisk2022} and Riemannian Geometry methods using spatial
covariances as features~\cite{Aydarkhanov2020} have both been shown
to perform well in the presence of \ac{erp} jitter.
In this work, we opted to apply \ac{cble} because it has successfully been applied to
classify jittered \acs{erp}.
We adapt the \ac{cble} to an iterative method akin to Woody's scheme that can better
compensate for jitter, to improve covert VSA decoding performance.

@@

Consider a training set of $N$ EEG epochs with $C$ channels of $S$
samples $\{\mat{X}_n^\mathrm{train} \in \mathbb{R}^{C\times S}\}_{n=1}^N$
with corresponding training labels $\mathbf{l^\mathrm{train}} \in \{\mathrm{target},
	\textrm{non-target}\}_{n=1}^N$, and a
similar testing set of $M$ epochs $\{\mat{X}_m^\mathrm{test} \in
	\mathbb{R}^{C\times S}\}_{m=1}^M$.
We assume that the sampling period is $T$, i.e., that the sample with index $s$ is sampled at time $sT$.
In the following, we use the matrix slicing notation to denote row or column intervals extracted from a matrix.
For instance, $\mat{X}[:,s_1:s_2]$ denotes all columns of $\mat{X}$ with indices between $s_1$ (included) and $s_2$ (excluded).

\Ac{cble}, summarized in \cref{sec:wcble/app/cble-alg}, works by training a
\textit{first-stage} classifier $\mathcal{C}(\theta,f)$
defined within a time period $[s_1:s_2]$ with a set of parameters $\theta$ and a
decision function $f(\mat{X}[:,s_1:s_2],\theta) \rightarrow
y$ outputting a classification score $y\in\mathbb{R}$ for a given epoch $\mat{X}$,
such that
\begin{equation}
  \theta = \mathrm{train}_\mathcal{C}(\{\mat{X}_n^\mathrm{train}[:,s_1:s_2]\}_{n=1}^N,\mathbf{l}^\mathrm{train})
  \label{eq:cble-train}
\end{equation}
Then,
$f$ can be applied to all (possibly overlapping) slices of length $s_2-s_1$ of
an epoch $\mat{X}$, resulting in a vector of score values
$\mathbf{y}=[y_1\,\ldots\,y_R]^T \in\mathbb{R}^R$ such that
\begin{equation}
  y_s = f(\mat{X}[:,s:s+(s_2-s_1)],\theta)\quad \forall s\in 1,\ldots,R
	\label{eq:score}
\end{equation}
with $R = S-(s_2-s_1)$.
To leverage \ac{cble} for \ac{erp} classification, the score vectors $\mathbf{y}$ can be
arranged in matrices $\mat{Y}^\mathrm{train}\in\mathbb{R}^{N\times R}$ and $\mat{Y}^\mathrm{test}\in\mathbb{R}^{M\times R}$.
These can be further classified by training a \textit{second-stage} classifier on
$\mat{Y}^\mathrm{train}$ and class labels $\mathbf{l}^\mathrm{train}$.
However, the resulting score-over-time vectors per epoch still suffer from jitter.
For classification, we follow the approach of \cite{Mowla2017}, using a
maximum-level hierarchical Daubechies-4 wavelet transform to reduce
dimensionality before classification with the second-stage classifier.
In the \ac{cble}-decoder, it is this wavelet transform that decreases the sensitivity
to latency differences, actively compensating for \ac{erp} latency jitter.

When using a simple spatiotemporal linear classifier as first-stage classifier,
\ac{cble} is equivalent to the first iteration of Woody's algorithm with the
spatiotemporal classifier weights as template.
\textcite{Thompson2012}, \textcite{Mowla2017}, and \textcite{Mowla2020} show
that \ac{cble} is relatively independent of the first-stage classifier for \ac{bci}
accuracy prediction and for \ac{erp} classification.
Therefore, we opt to use the variant of Linear Discriminant Analysis with block-Toeplitz
regularized covariance matrix (tLDA) proposed by \textcite{Sosulski2022}, as the first-stage
classifier and logistic regression as second stage.

\subsection{Robust latency features}
\label{sec:robust-latency}
The \ac{erp} decoding method based on \ac{cble} introduced by \textcite{Mowla2017} does
not make use of the extracted latencies, only passing score matrix $\mat{Y}$ on
to the second-stage classifier.
Furthermore, previous \ac{cble} works~\cite{Thompson2012, Mowla2020} only used these
latencies to correlate them with neurophysiological processes or to predict
decoder performance.
We noticed that, while \ac{cble} performance was unaffected, the classification
performance of our proposed method can be improved if the estimated
latencies are also made available as features to the second-stage classifier,
after a square transform for linear separability~\cite{Thompson2012}.

However, including these latency features gives rise to the following issue when
classifying unseen data.
The \ac{cble} latency estimate is defined only for target epochs as
$s_\mathrm{target}=\argmax_s{y_s}$.
This is the point in time where the target class reaches the largest separation
from the background noise and non-target class, indicating the target \ac{erp} is
most likely to occur here.
However, in a classifier test phase, it is not known a priori whether an
unseen epoch is a target or a non-target epoch.
This problem is solved by defining an estimated latency per class
$s_\mathrm{target}$ and $s_\mathrm{non-target}$ for every epoch,
regardless of its actual class.
The estimated class latencies can then be used as features for training and
testing the second-stage classifier.
This way, the latencies of the testing data can be presented to the second-stage
classifier without knowledge of the testing data class labels, making them
useful in decoding.

In a similar manner to the target latency, the non-target latency could be defined
as~$s_\mathrm{non-target}=\argmin_s{y_s}$.
However, this is problematic since it is not evident how to estimate
the latency of, e.g., a P3 \ac{erp} component for a non-target epoch, since the
non-target class is characterized by the absence of this component.
\footnote{Depending on the stimulation and experimental design, \ac{erp} components can be present in one experimental
	condition or class and missing in another, or appear in multiple classes
	at different amplitudes. In the latter case, it could be possible
	to estimate the latency of an \ac{erp} in the conditions where it appears, but if
	its amplitude is lower or negligible in some conditions, latency estimates
	will be less accurate, hence this case is still problematic.}.
In fact, $y$ can have multiple local minima or entirely lack distinct peaks for
non-targets, rendering the minimum estimate meaningless.
